Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-25 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revisions on abstract and introduction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2897 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 13d860415908960772a21bb81941abb8ebaa64eb

\title{Circular Distributional Trees and Forests with an Application to 
    Wind Direction Forecasting}
\Shorttitle{Circular Trees and Forests}

@@

\Abstract{
As circular data appears in many different areas, the development of
appropriate statistical models which can account for periodicity has become an
important but challenging research field. For probabilistic modeling of
circular data typically the von Mises distribution is employed. In order to
regress its parameters on covariates, a circular distributional tree is
proposed as an alternative to more commonly-used additive models. The resulting
circular trees are easy to interpret, can detect non-additive effects, and
select covariates and their interactions automatically. By combining an
ensemble of circular trees a more stable and regularized circular
distributional forest can be built which allows for modeling smooth effects as
well.

To evaluate the proposed circular trees and forests, a case study on
probabilistic forecasting of wind direction for two airports in Austria is
presented. In addition to the tree-based models, alternative common approaches
are used as a benchmark in order to assess the predictive performances of
circular trees and forests. 
}
\Keywords{Distributional Trees and Forests; Circular Response; von Mises Distribution}

@@

In many approaches to model circular data it is assumed that the circular
variable of interest follows a circular distribution, in particular the von
Mises distribution which is also known as the ``circular normal distribution''
(see Section~\ref{sec:prob_circ} for more details). This is also the basis
which early regression models for circular data are built on as they regress
the parameter(s) of the von Mises distribution on covariates. One of the first
regression models with a circular response variable and linear covariates was
presented by \cite{Gould:1969} where the circular mean, i.e., the location
parameter of the von Mises distribution is predicted by a linear combination of
covariates. \cite{Johnson+Wehrly:1978} took this idea one step further by
plugging in a link function such that the values of the linear predictor are
transformed to a restricted interval of length $2\pi$. This generalized linear
model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
subsequently by \cite{Fisher:1993} introducing another specific transformation
function which maps the values of the real line to the unit circle. Moreover,
the presented methodology allows for both distribution parameters, the mean
direction and the concentration parameter of the von Mises distribution, to be
modeled depending on covariates, not only in separate models but also in a
mixed model. While being built on well elaborated theory this methodology has
shown to be very challenging in its application, mainly due to the complexity
encountered in optimizing the log-likelihood function which is not globally
concave. Therefore, highly informative starting values are crucial for the
model to converge at all \citep{Pewsey+Neuhaeuser+Ruxton:2013,
Gill+Hangartner:2010}. In order to avoid this strong dependence on appropriate
initial values, \cite{Mulder+Klugkist:2017} presented an alternative approach
to regression modeling of circular data by employing a GLM in a Bayesian
framework. All listed methods and most of their further developments
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive
regression models. While these generally provide a powerful toolbox for
modeling particularly smooth effects, they may not be able to capture all types
of effects adequately. Especially abrupt changes might not be detected
properly owing to the enforced additive structure and model assumptions, such
as a possibly highly non-linear link function and the form of linear
combinations.

An alternative modeling approach that can deal with non-additive effects is
provided by tree-based methods. \cite{Lund:2002} introduced the concept of
circular regression trees which was further extended by
\cite{Larraondo+Inza+Lozano:2018}. Following this idea of employing tree-based
methods but also including the probabilistic concept of regressing distribution
parameters on covariates, we propose a very flexible approach by applying the
von Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular
data. The resulting circular distributional trees and forests do not require
any model assumptions as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure
while circular forests allow for modeling smooth changes as well. Thus, a high
level of flexibility is provided while a full distributional model is specified
offering a wide range of inference.

As a motivating example, a circular distributional tree is employed to
probabilistic wind direction forecasting. Wind direction is a classical
circular quantity and accurate forecasts are of great importance for
decision-making processes especially in air traffic management.
Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at the
airport of Innsbruck which lies at the bottom of a deep valley in the Alps.
Topography channels wind along the west-east valley axis or along a tributary
valley intersecting from the south. Hence, pressure gradients to which valley
wind regimes react are considered as covariates along with other meteorological
measurements at the airport (lagged by one hour), such as wind direction and
wind speed at the airport.

Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
empirical (gray) and fitted von Mises (red) distribution of wind direction in
each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5). (2) Downslope winds blowing across the Alpine crest
along the intersecting valley towards Innsbruck (from south-east to north-west,
node 8). (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be distinguished clearly into wind regimes
and consequently are associated with a very low estimated concentration $\hat
\kappa$. In terms of covariates, the lagged wind direction (``persistence'') is
mostly responsible for distinguishing the broad wind regimes listed above while
the pressure gradients and wind speed separate between subgroups with high vs.\
low precision. Note that in the meteorological context wind direction 
is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
%%
\begin{figure}[t]
  \centering
  \includegraphics[width = \textwidth]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
  \caption{Fitted tree based on the von Mises distribution for 1-hourly wind
  direction forecasts at the airport of Innsbruck. In each terminal node the
  empirical histogram (gray) and fitted density (red line) are depicted along
  with the estimated location parameter (red hand). The covariates employed are
  wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
  ($\text{dpressure; hPa}$) west, east and south of the airport, all lagged by
  one hour.}
  \label{fig:tree_ibk} 
\end{figure}
%%

A more extensive case study of circular distributional trees and forests
applied to probabilistic forecasting of wind direction at the airports of
Innsbruck and Vienna will be presented in this paper. Moreover, their
performances are benchmarked to alternative commonly used approaches. The
remainder of the paper is structured as follows: The theory on probabilistic
circular modeling introducing the von Mises distribution will be discussed in
Section~\ref{sec:prob_circ}. The methodology of circular distributional trees
and forests and their features are introduced in Section~\ref{sec:tree_forest}.
After the case study presented in \ref{sec:wind}, a comprehensive summary and
conclusion is given in Section~\ref{sec:summary}.

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
\newpage
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and commonly applied for probabilistic modeling of circular
data. Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in (-\pi, \pi]$ is given by:
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ \citep[see, e.g.,][for a more detailed
overview]{Jammalamadaka+Sengupta:2001}.

@@

Figure~\ref{fig:densities} depicts a distributional model fitted to circular
data employing the von Mises distribution. While the left panel only serves as
an illustration of the corresponding ``linear'' description restricted to an
interval of length $2\pi$ but consequently not accounting for periodicity, the
right panel displays an appropriate representation of a circular quantity on
the unit circle. Looking at the fitted density (red line) and estimated
location parameter (red hand) this example shows that reasonably well fitting
probabilistic models can be established for circular data by specifying an
appropriate distribution family such as the von Mises distribution and
estimating its parameters via maximum likelihood. However, this procedure
considers only the response variable. Including covariates can improve any
model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types
of data, including circular data.

% -------------------------------------------------------------------
% Circular distributional trees and forests
% -------------------------------------------------------------------
\section{Circular distributional trees and forests}
\label{sec:tree_forest}
A common way of including covariates is to regress the model parameters on
covariates. In the case of a probabilistic model for a circular response this
corresponds to expressing the location parameter $\mu$ and the concentration
parameter $\kappa$ of the von Mises distribution by means of the available
covariates, e.g., in the form of a linear combination, possibly with an
appropriate link function plugged in as in GLMs. This kind of approach is
presented for example in the mixed model of \cite{Fisher+Lee:1992} and
\cite{Fisher:1993}. Apart from the methodology developed specifically for
circular data more general approaches to regressing model parameters on
covariates are commonly used such as for example generalized additive models
for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}
where each distribution parameter is estimated by its own generalized additive
model (GAM). However, difficulties might arise in case of non-additive changes
and a proper model specification can be very challenging, particularly for a
high number of covariates and no information on possible interactions. As an
alternative regression approach which can deal with these problems, we
introduce a circular probabilistic tree-structured model in this section.

\subsection{Circular trees}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly
for complex data with strong variations. Therefore, separating the data set
into more homogeneous subgroups based on covariates before fitting a local
model in each of these subgroups allows to capture (possibly) group specific
effects more precisely and hence can result in an overall better fitting model.
This is the general idea of regression trees which are combined with
distributional modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the
unbiased recursive partitioning algorithms MOB
\citep{Zeileis+Hothorn+Hornik:2008} or CTree
\citep{Hothorn+Hornik+Zeileis:2006}. By applying the presented distributional
trees, a full distributional model is specified in each node of the tree for a
specific distribution family chosen in advance. Hence, selecting the von Mises
distribution allows for an application of distributional trees to circular data
following the tree building algorithm as introduced in Section~2.2. of
\cite{Schlosser+Hothorn+Stauffer:2019}.

In the applied tree-building algorithm, in each node the crucial decision on
how and where to split the data is based on model scores which are obtained by
evaluating the score function $s$ for the individual observations and parameter
estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
two distribution parameters and a data set of $n$ observations, this yields an
$n \times 2$ matrix that can be employed as a discrepancy measure, capturing
how well each given observation conforms with the estimated location
$\hat{\mu}$ and precision $\hat{\kappa}$, respectively. To capture dependence
on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test~(MOB) or a
permutation test~(CTree). By doing so in each partitioning step, the covariate
with the highest significant association (i.e., lowest significant $p$-value,
if any) is selected for splitting the data. The corresponding split point is
chosen either by optimizing the log-likelihood~(MOB) or a two-sample test
statistic~(CTree) over all possible partitions. This procedure is repeated
recursively until there are no significant parameter instabilities or until
another stopping criterion is met (e.g., subgroup size or tree depth).

@@

\subsection{Circular forests}
\label{sec:circforest} 
A natural extension of circular distributional trees are ensembles or forests
that can improve forecasts by regularizing and stabilizing the model. Random
forests introduced by \cite{Breiman:2001} average the predictions of an
ensemble of trees, each built on a subsample or bootstrap of the original data.
A generalization of this strategy is to obtain weighted predictions by adaptive
local likelihood estimation of the distributional parameters
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019,
Hothorn+Zeileis:2017}. More specifically, for each possibly new
observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights
$w_i^{forest}(\bm{z})$ is obtained that is based on the number of trees in
which $\bm{z}$ is assigned to the same terminal node as each learning
observation $y_i, i \in \{1,\ldots,n\}$. Hence, for a forest of $T$ trees the
weights are calculated via
%%
\begin{equation}
  w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
  \frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
%%
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree. Therefore, similar observations ending up more
often in the same terminal node have higher weights and in that way a stronger
influence in the estimation process.

@@

Therefore, the resulting parameter estimates can smoothly adapt to the given
covariates $\bm{z}$ whereas $w_i^{forest}(\bm{z}) = 1$ would correspond to the
unweighted full-sample estimates and $w_i^{forest}(\bm{z}) \in \{0, 1\}$
corresponds to the abrupt splits from a tree. Thus, circular distributional
forests can capture both smooth and abrupt changes, while covariates and
possible interactions are selected automatically allowing for an easy-to-apply
methodology with a full specification of the von Mises distribution.

% -------------------------------------------------------------------
% Case study
% -------------------------------------------------------------------
\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}
Accurate forecasts of wind directions are of great importance for risk
management in various fields such as agriculture, energy production or air
safety control. For example, in order to direct airplanes to a safe landing,
precise knowledge of wind direction for the next hour(s) at the respective
airport is highly desirable and adequate prediction methods are required. In
this section, a comprehensive case study on probabilistic wind direction
forecasting is provided for two airports in Austria with fundamentally
different site characteristics: Innsbruck with a rather unusual location for an
airport, located at the bottom of a deep Alpine valley, and Vienna located
within flat terrain in the East of Austria. In addition to the newly introduced
circular distributional trees and forests, the case study employs various
alternative approaches to forecast wind direction as a benchmark. The study is
based on $1$\,h and $3$\,h forecasts employing lagged observations in the
vicinity of the airports as predictor variables.

@@

    \citep{simon.etal:2017,stauffer.etal:2017}. Usually, the performance of

@@

    \citet{vogel.etal:2018} and is discussed in a comprehensive summary on
    different time-adaptive training schemes in \citet{lang.etal:2019c}.

@@

  \item \emph{Circular distributional tree:}

>>>>

\title{Circular Regression Trees and Forests with an Application to
Probabilistic Wind Direction Forecasting}
\Shorttitle{Circular Regression Trees and Forests}

@@

\Abstract{
Establishing a proper regression model for circular data such as animal movement
directions, periodic measurements of neuronal activity or wind directions has shown
to be very challenging. Commonly applied generalized linear and additive models 
can provide a useful toolbox, however, non-additive effects might not be detected 
and difficulties may arise in modeling smooth effects due to highly non-linear link 
functions and increasing irregularities when moving closer to the ``origin point'' 
(0 which is equal to $2\pi$ in the radiant case).
%% FIXME: clear explanation of difficulties
Therefore, we propose a new approach by applying distributional trees and 
forests to circular data. In particular, the parameters of the employed von Mises 
distribution are regressed on covariates by a tree-based model yielding a constant set 
of estimated distribution parameters for each resulting subgroup, regardless of any 
``circular shift''/the ``circular position''.
%As circular data appears in many different areas, the development of
%appropriate statistical models which can account for periodicity has become an
%important but challenging research field. For probabilistic modeling of
%circular data typically the von Mises distribution is employed. In order to
%regress its parameters on covariates, a circular regression tree is
%proposed as an alternative to more commonly-used generalized linear and additive 
%models. 
The resulting circular regression trees are easy to interpret, can detect non-additive effects, 
and select covariates and their interactions automatically. By combining an
ensemble of circular regression trees a more stable and regularized circular
regression forest can be built which allows for modeling smooth effects as
well.

To evaluate the proposed circular regression trees and forests, a case study on
probabilistic forecasting of wind direction for two airports in Austria is
presented. In addition to the tree-based models, alternative common approaches
are used as a benchmark in order to assess the predictive performances of
circular regression trees and forests. 
}
\Keywords{Distributional Trees, Random Forests; Circular Data; von Mises Distribution}

@@

In many approaches to model circular data it is assumed that the circular
variable of interest follows a circular distribution, in particular the von
Mises distribution which is also known as the ``circular normal distribution''
(see Section~\ref{sec:prob_circ} for more details). This is also the basis
which early regression models for circular data are built on as they regress
the parameter(s) of the von Mises distribution on covariates. One of the first
regression models with a circular response variable and linear covariates was
presented by \cite{Gould:1969} where the circular mean, i.e., the location
parameter of the von Mises distribution is predicted by a linear combination of
covariates. \cite{Johnson+Wehrly:1978} took this idea one step further by
plugging in a link function such that the values of the linear predictor are
transformed to a restricted interval of length $2\pi$. This generalized linear
model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
subsequently by \cite{Fisher:1993} introducing another specific transformation
function which maps the values of the real line to the unit circle. 
Moreover, the presented methodology does not only allow for the mean direction to be modeled 
depending on covariates while a constant estimate is obtained for the concentration parameter 
but also for the vice versa version. For each of these two approaches a separate model is introduced, 
however, they are also combined in a joint model (referred to as ``mixed model'').
%Moreover, the presented methodology allows for both distribution parameters, the mean
%direction and the concentration parameter of the von Mises distribution, to be
%modeled depending on covariates, not only separately but also in a
%combined model (also referred to ``mixed model''). 
In the employed iterative optimization algorithm the two sets of regression coefficients used
to model $\mu$ and $\sigma$ respectively are updated in alternation until convergence.
%To obtain proper initial values it is recommended to apply two separate models 
%where either one of the parameters is fixed while the other is predicted employing covariates 
%in a GLM-type model.
While the combined model but also the two separate models are built on well elaborated theory 
the employed methodology has shown to be very challenging in its application, mainly due to the 
complexity encountered in optimizing the log-likelihood function which is not globally concave. 
Therefore, highly informative starting values are crucial for the model to converge at all 
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. 
In order to avoid this strong dependence on appropriate initial values, \cite{Mulder+Klugkist:2017} 
present an alternative approach to regression modeling of circular data by employing a GLM in a 
Bayesian framework, however, only the mean direction is regressed on covariates.
All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
are built on additive regression models.
Generally these provide a powerful toolbox for modeling particularly smooth effects,
however, in the case of circular data a possibly highly non-linear link function
can induce a more complex behavior of the modeled parameters as the location
parameter shifts closely towards the ``origin point'' $0$, also depending on the 
direction of the shift.
Moreover, not all types effects of may be modeled adequately. Especially abrupt changes 
might not be detected properly owing to the enforced additive structure and selected model 
specification such as the form of linear combinations.
%% FIXME: clear explanation of difficulties close to 0

In order to avoid these difficulties in probabilistic regression modeling of circular data 
(and also the requirement of initial values) we propose an alternative and very flexible 
approach by applying the von Mises distribution within the methodology of distributional 
trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests 
(for short: \emph{circular trees and forests}) 
do not require any model assumptions as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while 
circular forests allow for modeling smooth changes as well. Thus, a high level of 
flexibility is provided while a full distributional model is specified in each terminal 
node offering a wide range of inference. Moreover, due to the parameters being estimated 
as constants for each subgroup difficulties connected to ``circular shifts'' close to $0$ 
are avoided.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} have already introduced the concept of circular regression 
trees where binary splits are made based on a circular distance which is evaluated between 
the observed response values and the conditional mean direction of the circular response variable
and in that way serves as a measure of homogeneity within subgroups. 
%For a circular response $\theta$, covariate(s) $z$ and the conditional mean direction $\mu(z)$
%the employed circular distance is defined by $1-\cos(\theta - \mu(z)$.
%In each terminal node a predicted values is provided by the sample mean direction of the corresponding 
%subset of observations.
While this approach and also further extensions \citealp{Larraondo+Inza+Lozano:2018} allow not only for 
linear but also for a circular covariate, the tree splits are performed based on effects in the expected 
value only, i.e., only the location parameter $\mu$ is regressed on covariates.
%Furthermore, even though an estimate of dispersion in the form of total node heterogeneity is provided 
%along with the predicted mean value in each terminal node
Hence, the here proposed circular regression trees and forests can be seen as a further step in the development
of tree-based circular models by incorporating the framework of distributional modeling and in that way
considering splits in all distribution parameters and providing a fully specified distribution fitted 
in each terminal node.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%An alternative modeling approach that can deal with non-additive effects is
%provided by tree-based methods. \cite{Lund:2002} introduced the concept of
%circular regression trees which was further extended by
%\cite{Larraondo+Inza+Lozano:2018}. Following this idea of employing tree-based
%methods but also including the probabilistic concept of regressing distribution
%parameters on covariates, we propose a very flexible approach by applying the
%von Mises distribution within the methodology of distributional trees and
%distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular
%data. The resulting circular distributional trees and forests do not require
%any model assumptions as covariates and their possible interactions are
%selected automatically. Non-additive effects are captured by the tree structure
%while circular forests allow for modeling smooth changes as well. Thus, a high
%level of flexibility is provided while a full distributional model is specified
%offering a wide range of inference.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection*{Motivating example}
To provide a first impression of the presented methodology a circular regression 
tree is employed to probabilistic wind direction forecasting. Wind direction is a 
classical circular quantity and accurate forecasts are of great importance for
decision-making processes especially in air traffic management.
Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at the
airport of Innsbruck which lies at the bottom of a deep valley in the Alps.
Topography channels wind along the west-east valley axis or along a tributary
valley intersecting from the south. Hence, pressure gradients to which valley
wind regimes react are considered as covariates along with other meteorological
measurements at the airport (lagged by one hour), such as wind direction and
wind speed at the airport.
%%
\begin{figure}[p!]
  \centering
  \includegraphics[width=0.95\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
  \caption{Fitted tree based on the von Mises distribution for 1-hourly wind
  direction forecasts at the airport of Innsbruck. In each terminal node the
  empirical histogram (gray) and fitted density (red line) are depicted along
  with the estimated location parameter (red hand). The covariates employed are
  wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
  ($\text{dpressure; hPa}$) west, east and south of the airport, all lagged by
  one hour.}
  \label{fig:tree_ibk} 
\end{figure}
%%
Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
empirical (gray) and fitted von Mises (red) distribution of wind direction in
each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5). (2) Downslope winds blowing across the Alpine crest
along the intersecting valley towards Innsbruck (from south-east to north-west,
node~8). (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be distinguished clearly into wind regimes
and consequently are associated with a very low estimated concentration $\hat
\kappa$. In terms of covariates, the lagged wind direction (``persistence'') is
mostly responsible for distinguishing the broad wind regimes listed above while
the pressure gradients and wind speed separate between subgroups with high vs.\
low precision. Note that in the meteorological context wind direction 
is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).

A more extensive case study of circular regression trees and forests
applied to probabilistic forecasting of wind direction at the airports of
Innsbruck and Vienna will be presented in this paper. Moreover, their
performances are benchmarked to alternative commonly used approaches. The
remainder of the paper is structured as follows: The theory on probabilistic
circular modeling introducing the von Mises distribution will be discussed in
Section~\ref{sec:prob_circ}. The methodology of circular regression trees
and forests and their features are introduced in Section~\ref{sec:tree_forest}.
After the case study presented in \ref{sec:wind}, a comprehensive summary and
conclusion is given in Section~\ref{sec:summary}.

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
%\newpage
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and commonly applied for probabilistic modeling of circular
data. Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in (-\pi, \pi]$ is given by:
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ \citep[see, e.g.,][for a more detailed
overview]{Jammalamadaka+Sengupta:2001}.

@@

Figure~\ref{fig:densities} depicts a distributional model fitted to circular
data employing the von Mises distribution. While the left panel only serves as
an illustration of the corresponding ``linear'' description restricted to an
interval of length $2\pi$ but consequently not accounting for periodicity, the
right panel displays an appropriate representation of a circular quantity on
the unit circle. Looking at the fitted density (red line) and estimated
location parameter (red hand) this example shows that reasonably well fitting
probabilistic models can be established for circular data by specifying an
appropriate distribution family such as the von Mises distribution and
estimating its parameters via maximum likelihood. However, this procedure
considers only the response variable. Including covariates can improve any
model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types
of data, including circular data.
Already existing generalized linear or additive approaches to circular regression 
models employ linear combinations of the covariates to connect them to one (or more) 
distribution parameters. 
%For the mean direction~$\mu$ \cite{Gould:1969} proposes the model 
%\begin{equation}
%\mu = \mu_0 + \sum_j \beta_j\cdot x_j(\textit{mod } 2\pi)
%\end{equation}
%with a circular intercept $\mu_0$ and $x$ being a vector of covariates
%with the corresponding vector of coefficients $\beta$.
By following this idea and including a link function $g$ many other models
have been developed in the general form 
\begin{equation}
\mu = \mu_0 + g(\beta^{\top} x).
\end{equation}
For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
on an intervall of length $2\pi$ by setting $g(x) = 2\pi \cdot F(x)$ for a cumulative 
distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. The concentration parameter

%% FIXME: information on already existing generalized linear and additive approaches
%% with formulas (\mu = g(X \beta) + \mu_0)   FIXME: add mu_0?
%% clear description of difficulties

% -------------------------------------------------------------------
% Circular regression trees and forests
% -------------------------------------------------------------------
\section{Circular regression trees and forests}
\label{sec:tree_forest}

%% FIXME: what are we aiming at? which problems do we want to avoid/solve?
%% mu and kappa as constants in subgroups, no initial values
%% Lund: what do they present (including formulas)
%% compare, what is different, how do we improve this approach

A common way of including covariates is to regress the model parameters on
covariates. In the case of a probabilistic model for a circular response this
corresponds to expressing the location parameter $\mu$ and the concentration
parameter $\kappa$ of the von Mises distribution by means of the available
covariates, e.g., in the form of a linear combination, possibly with an
appropriate link function plugged in as in GLMs. This kind of approach is
presented for example in the mixed model of \cite{Fisher+Lee:1992} and
\cite{Fisher:1993}. Apart from the methodology developed specifically for
circular data more general approaches to regressing model parameters on
covariates are commonly used such as for example generalized additive models
for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}
where each distribution parameter is estimated by its own generalized additive
model (GAM). However, difficulties might arise in case of non-additive changes
and a proper model specification can be very challenging, particularly for a
high number of covariates and no information on possible interactions. As an
alternative regression approach which can deal with these problems, we
introduce a circular probabilistic tree-structured model in this section.

\subsection{Circular regression trees}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly
for complex data with strong variations. Therefore, separating the data set
into more homogeneous subgroups based on covariates before fitting a local
model in each of these subgroups allows to capture (possibly) group specific
effects more precisely and hence can result in an overall better fitting model.
This is the general idea of regression trees which are combined with
distributional modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the
unbiased recursive partitioning algorithms MOB
\citep{Zeileis+Hothorn+Hornik:2008} or CTree
\citep{Hothorn+Hornik+Zeileis:2006}. By applying the presented distributional
trees, a full distributional model is specified in each node of the tree for a
specific distribution family chosen in advance. Hence, selecting the von Mises
distribution allows for an application of distributional trees to circular data
following the tree building algorithm as introduced in Section~2.2. of
\cite{Schlosser+Hothorn+Stauffer:2019}.

In the applied tree-building algorithm, in each node the crucial decision on
how and where to split the data is based on model scores which are obtained by
evaluating the score function $s$ for the individual observations and parameter
estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
two distribution parameters and a data set of $n$ observations, this yields an
$n \times 2$ matrix that can be employed as a discrepancy measure, capturing
how well each given observation conforms with the estimated location
$\hat{\mu}$ and precision $\hat{\kappa}$, respectively. To capture dependence
on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test~(MOB) or a
permutation test~(CTree). By doing so in each partitioning step, the covariate
with the highest significant association (i.e., lowest significant $p$-value,
if any) is selected for splitting the data. The corresponding split point is
chosen either by optimizing the log-likelihood~(MOB) or a two-sample test
statistic~(CTree) over all possible partitions. This procedure is repeated
recursively until there are no significant parameter instabilities or until
another stopping criterion is met (e.g., subgroup size or tree depth).
A more detailed description of the applied tree-building algorithm,
particularly of CTree, can be found in Appendix~\ref{app:tree}.

@@

\subsection{Circular forests}
\label{sec:circforest} 
A natural extension of circular regression trees are ensembles or forests
that can improve forecasts by regularizing and stabilizing the model. Random
forests introduced by \cite{Breiman:2001} average the predictions of an
ensemble of trees, each built on a subsample or bootstrap of the original data.
A generalization of this strategy is to obtain weighted predictions by adaptive
local likelihood estimation of the distributional parameters
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019,
Hothorn+Zeileis:2017}. More specifically, for each possibly new
observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights
$w_i^{forest}(\bm{z})$ is obtained that is based on the number of trees in
which $\bm{z}$ is assigned to the same terminal node as each learning
observation $y_i, i \in \{1,\ldots,n\}$. Hence, for a forest of $T$ trees the
weights are calculated via
%%
\begin{equation}
  w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
  \frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
%%
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree. Therefore, similar observations ending up more
often in the same terminal node have higher weights and in that way a stronger
influence in the estimation process.

@@

Therefore, the resulting parameter estimates can smoothly adapt to the given
covariates $\bm{z}$ whereas $w_i^{forest}(\bm{z}) = 1$ would correspond to the
unweighted full-sample estimates and $w_i^{forest}(\bm{z}) \in \{0, 1\}$
corresponds to the abrupt splits from a tree. Thus, circular regression
forests can capture both smooth and abrupt changes, while covariates and
possible interactions are selected automatically allowing for an easy-to-apply
methodology with a full specification of the von Mises distribution.

% -------------------------------------------------------------------
% Case study
% -------------------------------------------------------------------
\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}
Accurate forecasts of wind directions are of great importance for risk
management in various fields such as agriculture, energy production or air
safety control. For example, in order to direct airplanes to a safe landing,
precise knowledge of wind direction for the next hour(s) at the respective
airport is highly desirable and adequate prediction methods are required. In
this section, a comprehensive case study on probabilistic wind direction
forecasting is provided for two airports in Austria with fundamentally
different site characteristics: Innsbruck with a rather unusual location for an
airport, located at the bottom of a deep Alpine valley, and Vienna located
within flat terrain in the East of Austria. In addition to the newly introduced
circular regression trees and forests, the case study employs various
alternative approaches to forecast wind direction as a benchmark. The study is
based on $1$\,h and $3$\,h forecasts employing lagged observations in the
vicinity of the airports as predictor variables.

@@

    \citep{Simon+Umlauf+Zeileis:2017, Stauffer+Mayr+Messner:2017}. Usually, the performance of

@@

    \citet{Vogel+Knippertz+Fink:2018} and is discussed in a comprehensive summary on
    different time-adaptive training schemes in \citet{Lang+Lerch+Mayr:2019}.

@@

  \item \emph{Circular regression tree:}
    For the circular regression tree introduced in Section~\ref{sec:circtree},
